---
title: "Homework1"
author: "Roman Krass, Keijo Nierula ..."
date: "2024-04-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, include=FALSE}

library(tidyverse)
library(tsibble)
library(slider)
library(fable)
library(fabletools)
library(forecast)
library(stats)
library(readxl)
library(feasts)
```

```{r importData, echo=FALSE, include=FALSE}

unemploy <- read_csv("./UNEMPLOY.csv")
population <- read_csv("./POP.csv")
personalSavingRate <- read_csv("./PSAVERT.csv")

```
## Part 1
```{r, echo=FALSE}

#head(unemploy, 3)
#head(population, 3)
#head(personalSavingRate, 3)
cbind(head(unemploy, 3), head(population, 3), head(personalSavingRate, 3))
```
When loooking into the different datasets, we can see that the data doesn't have the same time period. Because of that when combining the three datasets there would be some NAs if we just would join the data. We can solve this by using the smallest common date range of the three datasets.

```{r, echo=FALSE}
# Combine the three datasets to one tsibble with the DATE starting from 1959-01-01
data <- unemploy %>% 
    filter(DATE >= "1959-01-01") %>% 
  left_join(population, by = "DATE") %>% 
  left_join(personalSavingRate, by = "DATE") %>% 
  rename(unemployment = UNEMPLOY, population = POP, savingRate = PSAVERT) %>% 
  as_tsibble(index = DATE)

```

```{r timePlot Unemployment, out.width="80%", echo=FALSE}

# Plot the unemployment over time
ggplot(data, aes(x = DATE)) +
  geom_line(aes(y = unemployment), color = "red") +
    labs(title = "Unemployment rate over time",
        y = "Unemployment in thousands of people",
        x = "Year") +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
     theme_minimal()

```

In the plot we can see that the number of unemployed people has a lot of variation over time. But in this plot we just see the number of people that are unemployed. It would be interesting to see the unemployment rate in percent.
What also can be seen very clearly is the COVID-19 pandemic in 2020. The number of unemployed people increased a lot in this year.
In the plor we can also see an increase in the number of unemployed people in the years 2008 and 2009. This is the financial crisis that happened in these years.

```{r timePlot SavingRate, out.width="80%", warning=FALSE, echo=FALSE}
# Plot the personal saving rate over time
ggplot(data, aes(x = DATE)) +
  geom_line(aes(y = savingRate), color = "red") +
    labs(title = "Personal saving rate in % over time",
        y = "Personal saving in %",
        x = "Year") +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    scale_y_continuous(breaks = seq(0, , 35, by = 2)) +
    theme_minimal()

```

In the plot we can see that the personal saving rate decreased from about 1976 to 2007 from about 13% to 2%. After that the saving rate increased again to about 7% in 2020.
Something extreme in the plot is the increase of the saving rate during the COVID-19 pandemic from 2019 to 2022. The saving rate increased from about 7% to 32% in 2020. This is a huge increase in the saving rate. After that the saving rate had a big de and increase. After the pandemic the saving rate dropped to about 3%. This makes sense because the people saved a lot of money during the pandemic and after the pandemic they started to spend the money again.

## Part 2
We would like to visualize how the ruling party and the election year affect unemployment and personal
saving rate. Recreate the timeplots with the ruling parties shown by their color. Indicate also the election
years on your plots. Start from Carter’s presidency.

```{r, echo=FALSE, warning=FALSE}


# Create a sequence of dates for each presidency, assign the corresponding party, and merge with the data
dataPresidential <- data %>%
  left_join(presidential %>%
              mutate(DATE = map2(start, end, seq, by = "day")) %>%
              unnest(DATE) %>%
              select(DATE, party), by = "DATE") %>%
  filter(DATE >= as.Date("1977-01-20"))

# Add a new grouping variable
dataPresidential <- dataPresidential %>%
  mutate(party_group = cumsum(party != lag(party, default = party[1])))

# Plot the data
ggplot(dataPresidential, aes(x = DATE)) +
  geom_line(aes(y = savingRate, color = party, group = party_group), size = 1) +
   geom_vline(data = presidential, aes(xintercept = as.numeric(start)), linetype = "dashed", color = "black") +
  scale_color_manual(values = c("Republican" = "red", "Democratic" = "blue")) +
  labs(title = "Personal saving rate in % over time",
       y = "Personal saving in %",
       x = "Year",
       color = "Party") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_y_continuous(breaks = seq(0, 35, by = 2)) +
  theme_minimal()

```


```{r, echo=FALSE, warning=FALSE}

# Plot the data
ggplot(dataPresidential, aes(x = DATE)) +
  geom_line(aes(y = unemployment, color = party, group = party_group), size = 1) +
  geom_vline(data = presidential, aes(xintercept = as.numeric(start)), linetype = "dashed", color = "black") +
  scale_color_manual(values = c("Republican" = "red", "Democratic" = "blue")) +
  labs(title = "Unemployment rate over time",
       y = "Unemployment in thousands of people",
       x = "Year",
       color = "Party") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  theme_minimal()

```

## Part 3
Decompose the personal saving rate series using a centered moving average. Graph the season-adjusted
data. Explain whether the remainder is a white noise. If you believe that a transformation could be useful
before decomposing, then please do it. Compare your decomposition to the STL decomposition.
```{r, echo=FALSE, warning=FALSE}

dataDecomposed <- data %>%
  mutate('5-MA' = slider::slide_dbl(savingRate, mean, .before = 2, .after = 2, .complete = TRUE)) %>%
  mutate('11-MA' = slider::slide_dbl(savingRate, mean, .before = 5, .after = 5, .complete = TRUE))

ggplot(dataDecomposed, aes(x = DATE)) +
  geom_line(aes(y = savingRate, color = "Saving Rate")) +
  geom_line(aes(y = `5-MA`, color = "5-MA")) +
  geom_line(aes(y = `11-MA`, color = "11-MA")) +
  labs(title = "Personal saving rate in % over time",
       y = "Personal saving in %",
       x = "Year") +
  scale_color_manual(values = c("Saving Rate" = "black", "5-MA" = "red", "11-MA" = "blue")) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_y_continuous(breaks = seq(0, 35, by = 2)) +
  theme_minimal()


```


```{r, echo=FALSE, warning=FALSE}
#Graph the season-adjusted data.

data |>
  mutate(DATE = yearmonth(DATE)) |>
  as_tsibble(index = DATE, key = c(unemployment, population, savingRate), regular = TRUE) -> data_tsibble



# check if data$savingRate has na values
sum(is.na(data$savingRate))

# remove that one NA value
data <- data %>% drop_na()

# Convert savingRate to a ts object with a specified frequency
savingRate_ts <- ts(data$savingRate, frequency = 12)

# Decompose the savingRate time series with stl
stl_result <- stl(savingRate_ts, t.window = 12, s.window = "periodic", robust = TRUE)
autoplot(stl_result)

# Create a data frame with the original and adjusted savingRate
adjusted_data <- data %>%
  mutate(AdjustedSavingRate = savingRate - stl_result$time.series[, "seasonal"])

# Plot the original and adjusted savingRate
ggplot(adjusted_data, aes(x = DATE)) +
  geom_line(aes(y = savingRate, color = "Original"), ) +
  geom_line(aes(y = AdjustedSavingRate, color = "Seasonal adjusted"), ) +
  labs(title = "Saving Rate seasonal adjusted over time",
       y = "Saving Rate",
       x = "Year") +
  scale_color_manual(values = c("Original" = "red", "Seasonal adjusted" = "blue"))

```


# Part 4

Data Sources we used:

* Unemployment per canton: https://www.bfs.admin.ch/bfs/de/home/statistiken/arbeit-erwerb/erwerbslosigkeit-unterbeschaeftigung/registrierte-arbeitslose-seco.assetdetail.30245369.html
* Population statistics: https://www.bfs.admin.ch/bfs/de/home/statistiken/bevoelkerung/stand-entwicklung.assetdetail.26565154.html

Both sheets have been manipulated manually to make the parsing in R easier, e.g. we removed unnecessary columns and rows and for the population dataset we merged all the values from the years into one sheet.
Population data for 2023 was not yet available, which is why the combined tsibble has some NA values.
```{r, echo=FALSE, warning=FALSE}
unemployment <- read_xlsx("unemployment_rate_switzerland.xlsx")
population <- read_xlsx("population_numbers_switzerland.xlsx")

# Data Processing to make it possible to create a tsibble
unemployment <- unemployment %>%
  pivot_longer(cols = -Kanton, names_to = "Year", values_to = "Unemployment") %>%
  mutate(Year = as.numeric(Year))
population <- population %>%
  pivot_longer(cols = -Kanton, names_to = "Year", values_to = "Population") %>% mutate(Year = as.numeric(Year))

# Convert the data to a tsibble
combined_data <- full_join(unemployment, population, by = c("Year", "Kanton"))
combined_tsibble <- combined_data %>%
  as_tsibble(index = Year, key = Kanton)

# Write the data to a rds file
write_rds(combined_tsibble, "./task4.rds")

# Decompose the unemployment rate by canton for the plot
decomposed <- combined_tsibble %>%
  model(STL(Unemployment ~ season(window = "periodic") + trend(window = 12))) %>%
  components()

# Plot the seasonal and trend components of the unemployment rate by canton
ggplot(decomposed, aes(x = Year)) +
  geom_line(aes(y = season_adjust), color = "blue") +
  geom_line(aes(y = trend), color = "red") +
  facet_wrap(~ Kanton, scales = "free_y") +
  labs(x = "Year", y = "Unemployment Rate", color = "Component", title = "Unemployment Rate in Swiss Cantons") +
  theme_minimal() +
  theme(strip.text = element_text(size = 6), axis.text.x = element_text(size = 6), axis.text.y = element_text(size = 6))
```

The plot shows the seasonal and trend components of the unemployment rate in Switzerland by canton. The blue line represents the seasonal component, while the red line represents the trend component. With the seasonal line, peaks in unemployment, like the Covid-19 Pandemic can be seen clearly.
Also from the Task Description it was not clear for us whether the additional data (population data) also has to be displayed in the plot. We decided not do it, as it would have made the plot unreadable.